---
title: "Exploring the Determinants of Biogas Generation Potential in California"
author: "Jibikeoluwa Faborode, Yinan Ding, Abhay Rao"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
subtitle: Web address for GitHub repository
geometry: margin=2.54cm
fontsize: 12pt
mainfont: Times New Roman
---

```{=tex}
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 


\newpage
```
```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(sf)
library(mapview)
library(RColorBrewer)
library(cowplot)

# Set your ggplot theme
mytheme <- theme_gray(base_size = 10) +
  theme(axis.text = element_text(color = "darkgrey"), 
        legend.position = "bottom")

theme_set(mytheme)

# Load your datasets
biogas_sf<- st_read('../Data/biomethane/Biomethane.shp')
svi2018_raw <- read.csv('https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/California_COUNTY.csv')

```
# Table of figures

Figure 1: Population density across counties im California

Figure 2: Impoverished population density across counties in California

Figure 3: Total Methane potential across counties in California

Figure 4: Waste water Methane Potential across counties in California

Figure 5: Landfill methane potential across counties in California

Figure 6: Organic Waste methane potential across counties in California

Figure 7: Animal manure methane potential across counties in California

Figure 8: 

Figure 9:

Figure 10:

Figure 11:

Figure 12:

Figure 13:

Figure 14:

\newpage

# Background

## Rationale of the study
The transition to cleaner and renewable sources of energy requires more work to 
scale already existing gains, for example in the area of biogas utilization, 
while also seeking out new opportunities. Biogas is a form of renewable energy 
produced by anaerobic decomposition or thermochemical conversion of biomass such
as agricultural waste, manure, municipal waste sewage, green waste and food 
waste. Biogas is composed mostly of methane, alongside carbon dioxide, water 
vapor and other gases. Similar to natural gas, biogas can be burned directly as 
a fuel or treated to remove the CO2 and other gases before for being used in the
form of biomethane.

Utilizing biogas as an energy source helps to transform harmful gases from 
decomposing waste into positive use. Methane is a powerful greenhouse gas that 
traps heat in the atmosphere, with a global warming potential estimated to be 
over 25 times as potent as that of Carbon Dioxide.Transforming waste into biogas
therefore reduces greenhouse gas emissions and the risk of pollution to 
waterways. The [UNECE](https://unece.org/challenge#:~:text=Methane%20is%20a%20powerful%20greenhouses,grows%20to%2084%2D86%20times.) estimates that over a 20 year period, this ratio 
increases to 84-86 times. However, stored biogas limits the amount of methane 
released into the atmosphere and reduces dependence on fossil fuels. When stored
, biogas serves as a renewable and reliable baseload power source and can even 
be used to rapidly meet peak power demands. When biogas is used for energy 
generation in place of fossil fuels, it enables even more  emission reductions, 
sometimes resulting in carbon negative systems. According to the 
[Environmental and Energy Study Institute, EESI](https://www.eesi.org/papers/view/fact-sheet-biogasconverting-waste-to-energy)
,the reduction of methane emissions derived from tapping all the potential 
biogas in the United States would be equal to the annual emissions of 800,000 to
11 million passenger vehicles.

Unfortunately, despite the various benefits offered by biogas energy, the United
States currently only has 2,200 operating biogas systems. The EESI estimates 
this current capacity to be less than 20% of the total potential. It is 
on this basis that this study attempts to examine the factors contributing to 
biogas generation potential, particularly in California. California was selected
because of its status as a leading state in biogas generation potential based on
estimates by the National Renewable Energy Laboratory (NREL). According to the
[American Biogas Council](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/viewer.html?pdfurl=https%3A%2F%2Famericanbiogascouncil.org%2Fwp-content%2Fuploads%2F2019%2F05%2FABCBiogasStateProfile_CA.pdf&clen=226298&chunk=true), California can power nearly 200,000 homes if this biogas 
is utilized appropriately. The results of our analysis for California can 
provide the basis for further analysis across several other US states.



## Research Questions
The analysis conducted was done based on two research questions: 

1. Is biogas generation potential correlated with population density across 
California counties?

2. What additional factors must be considered in determining biogas generation 
potential?



\newpage

# Dataset Information
the datasets used include the following:

1. Methane Generation Potential
The NREL's dataset on methane generation potential covers data for the entire 
United States of America. It particularly providing information 
for Methane  generation potential across all states in metric tons from sources 
which include landfills, industrial organic waste, animal manure, and wastewater
. All of these individual sources were aggregated to estimate total methane 
generation potential by state. The dataset contained estimates for 2009 to 2012,
being the latest available data we could find online.

2. Population data 
Population data was pulled from the latest available the Social Vulnerability 
Index data compiled by the Center for Disease Control (CDC) and Agency for Toxic
Substance and Diseace Reegistry (ATSDR). This dataset also includes details on 
impoverished populations and spans from 2017-2018. Since the SDI offers data 
that helps to effectly plan towards meeting the needs of improverished 
populations, this dataset also helps to dds another dimension to our analysis as
it enables the exploration of impacts on socially vulnerable people.


\newpage

# Exploratory Analysis

The two datasets were wrangled and joined both datasets to have a unified 
dataframe that captured data on methane generation potential, population and 
poverty levels by county. An initial exploration of the data was done to provide
a visual assessment of spatial data and determine the emergence of any any 
trends.

```{r include=FALSE, echo=FALSE, warning=FALSE}
#Wrangling datasets
biogas_california_sf<- st_read('../Data/biomethane/Biomethane.shp') %>%
  filter(STATE_NAME == "California") #NREL Data for California
 
svi2018_CA_raw <- read.csv('https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/California_COUNTY.csv',
  colClasses = c('FIPS' = 'factor')) %>% 
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV, E_MINRTY)
# Population and SVI data for California

svi_sf_join <- biogas_california_sf %>% 
  left_join(svi2018_CA_raw, by = c("FIPS" = "FIPS") )
# Joined data.

str(svi_sf_join) # Columns of interest are OWCH4T, AMCH4T, WWTPCH4t, LFGCH4T, 
# E_TOTPOP, E_POV

# Change column names to make them explanatory
#colnames(svi_sf_join)[which(names(svi_sf_join) == "E_TOTPOP")] <- "Total Population"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "E_POV")] <- "Vulnerable Population"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "TotalCH4t")] <- "Total Methane Potential"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "WWTPCH4t")] <- "Wastewater Methane Potential"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "LFGCH4t")] <- "Landfill Methane Potential"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "OWCH4t")] <- "Organic Waste Methane Potential"
#colnames(svi_sf_join)[which(names(svi_sf_join) == "AMCH4t")] <- "Animal Waste Methane Potential"


```

```{r echo=FALSE, fig.height=3, fig.width=8, message=FALSE, warning=FALSE}
head(svi_sf_join) #Show the head of the new dataframe which includes population
```

Visually, the most populated area of California, given by the county shaded 
yellow, Los Angeles, has the highest methane (CH4) generation potential. The density of 
impoverished population appears to be similarly distributed.

```{r echo=FALSE, warning=FALSE}
Total_pop_plot <-ggplot(data=svi_sf_join) +
  geom_sf(aes(fill= E_TOTPOP)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 1: Population Density, CA', fill = "Total Population") + 
  theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  

Total_pov_plot <-ggplot(data=svi_sf_join) +
  geom_sf(aes(fill=E_POV)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 2: Impoverished Population Density, CA', 
       fill = "Impoverished Population") + 
  theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))

 Combinedplot_pop <- plot_grid(Total_pop_plot,Total_pov_plot)      
  print(Combinedplot_pop)
  
```
The total Methane generation potential is greatest at the most densely populated
areas. This trend appears to continue with wastewater derived methane.

```{r echo=FALSE, warning=FALSE}
Total_ch4_plot <-ggplot(data=svi_sf_join) +
  geom_sf(aes(fill= TotalCH4t)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 3: Total Methane Potential, CA (MT/year)', 
       fill = "Total Methane Potential") +
           theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  

WW_ch4_plot <- ggplot(data=svi_sf_join) +
  geom_sf(aes(fill=WWTPCH4t)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 4: Wastewater Methane Potential, CA (MT/year)', 
       fill = "Wastewater Methane Potential") +
           theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  

Combinedplot_totalww <- plot_grid(Total_ch4_plot,WW_ch4_plot)      
  print(Combinedplot_totalww)
```

Industrial Organic waste, like wastewater, is highly correlated with population.With landfill-based methane, some counties in central CA, which do not appear to have a high population, have comparatively high methane generation potential. There may be some non-population related factors in play.
```{r echo=FALSE, warning=FALSE}
Landfill_ch4_plot <- ggplot(data=svi_sf_join) +
  geom_sf(aes(fill=LFGCH4t)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 5: Landfill Methane Potential, CA (MT/year)',  
       fill = "Landfill Methane Potential") +
           theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  

OW_ch4_plot <- ggplot(data=svi_sf_join) +
  geom_sf(aes(fill=OWCH4t)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 6: Organic Waste Methane Potential, CA (MT/year)',  
       fill = "Organic Waste Methane Potential") +
           theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  

Combinedplot_lfow <- plot_grid(Landfill_ch4_plot,OW_ch4_plot)      
  print(Combinedplot_lfow)

```

The trend with animal-manure derived CH4 however substantially differs from the 
previous cases. Population does not appear to be a decisive factor.


```{r echo=FALSE, warning=FALSE}
AM_ch4_plot <- ggplot(data=svi_sf_join) +
  geom_sf(aes(fill=AMCH4t)) + 
  scale_fill_continuous(type='viridis')+
  labs(title='Figure 7: Animal Manure Derived Methane Potential, CA (MT per year)',
  fill = "Organic Waste Methane Potential") +
           theme(legend.title = element_text(color = "black", size = 8),
  legend.text = element_text(color = "black", size = 4), plot.title = 
    element_text(color = "black", size = 10, 
                            face = "bold", hjust = 0.5))
  
print(AM_ch4_plot)

```

\newpage

# Analysis

## Question 1: What factors affect biogas generation potential in different counties of California? Could it have potential correlations with the population size?

```{r echo=FALSE, message=FALSE, warning=FALSE}
mytheme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") 

cor.test(svi_sf_join$TotalCH4t,svi_sf_join$E_TOTPOP)
#cor:0.819
regression_pop_CH4<- lm(svi_sf_join$TotalCH4t~svi_sf_join$E_TOTPOP)
summary(regression_pop_CH4)
#p-value:4.22e-15
#Residual standard error: 20420 on 56 degrees of freedom
#Multiple R-squared: 0.6701
plot(regression_pop_CH4)

Plot_pop_CH4 <- 
  ggplot(svi_sf_join, aes(x = TotalCH4t, y = E_TOTPOP)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 5.0e+06)+
  labs(y= "Population Density", x = "Biogas Generation Potential")+
  ggtitle("Biogas Generation Potential v.s. Population in California")+
  mytheme
  print(Plot_pop_CH4)  

OW_Plot <- 
  ggplot(svi_sf_join, aes(x = OWCH4t, y = E_TOTPOP)) +
  geom_point(alpha=2, color = "Red") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 5.0e+06)+
  labs(y= "Population Density", x = "Organic Waste")+
  ggtitle("Organic Waste Potential")+
  mytheme

cor.test(svi_sf_join$OWCH4t,svi_sf_join$E_TOTPOP)
#cor: 0.96
regression_pop_OW<- lm(svi_sf_join$OWCH4t~svi_sf_join$E_TOTPOP)
summary(regression_pop_OW)
#p-value: < 2.2e-16
#Residual standard error: 515.6 on 56 degrees of freedom
#Multiple R-squared: 0.99
  
AM_Plot <- 
  ggplot(svi_sf_join, aes(x = AMCH4t, y = E_TOTPOP)) +
  geom_point(alpha=2, color = "Blue") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 5.0e+06)+
  labs(y= "Population Density", x = "Animal Manure")+
  ggtitle("Animal Manure Potential")+
  mytheme
  
cor.test(svi_sf_join$AMCH4t,svi_sf_join$E_TOTPOP)
#cor: -0.0018
regression_pop_AM<- lm(svi_sf_join$AMCH4t~svi_sf_join$E_TOTPOP)
summary(regression_pop_AM)
#p-value: < 0.989
#Residual standard error: 14190 on 56 degrees of freedom
#Multiple R-squared: -0.01785 
  
WWTP_Plot <- 
  ggplot(svi_sf_join, aes(x = WWTPCH4t, y = E_TOTPOP)) +
  geom_point(alpha=2, color = "Green") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 5.0e+06)+
  labs(y= "Population Density", x = "Wastewater")+
  ggtitle("Waste Water Potential")+
  mytheme

cor.test(svi_sf_join$WWTPCH4t,svi_sf_join$E_TOTPOP)
#cor: 0.99
regression_pop_WWTP<- lm(svi_sf_join$WWTPCH4t~svi_sf_join$E_TOTPOP)
summary(regression_pop_WWTP)
#p-value: < 2.2e-16
#Residual standard error: 1128 on 56 degrees of freedom
#Multiple R-squared:  0.984 
  
LF_Plot <- 
  ggplot(svi_sf_join, aes(x = LFGCH4t, y = E_TOTPOP)) +
  geom_point(alpha=2, color = "Blue") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 5.0e+06)+
  labs(y= "Population Density", x = "Landfill")+
  ggtitle("Landfill Potential")+
  mytheme
  
cor.test(svi_sf_join$LFGCH4t,svi_sf_join$E_TOTPOP)
#cor: 0.79
regression_pop_LF<- lm(svi_sf_join$LFGCH4t~svi_sf_join$E_TOTPOP)
summary(regression_pop_LF)
#p-value: < 1.926e-13
#Residual standard error: 11680 on 56 degrees of freedom
#Multiple R-squared:  0.6223 
Combinedplot <- plot_grid(OW_Plot,AM_Plot,WWTP_Plot,LF_Plot)
  print(Combinedplot)

## Question 2: Is there a relationship between methane generation potential and impoverished populations?

```{r message=FALSE, echo=FALSE, warning=FALSE}

#Correlation and regressions between methane and impoverished population

cor.test(svi_sf_join$TotalCH4t,svi_sf_join$E_POV)#0.82
regression_pov_CH4<- lm(svi_sf_join$TotalCH4t~svi_sf_join$E_POV)
summary(regression_pov_CH4) #p-value: 8.476e-16

cor.test(svi_sf_join$LFGCH4t,svi_sf_join$E_POV) #0.78
regression_pov_LFGCH4<- lm(svi_sf_join$LFGCH4t~svi_sf_join$E_POV)
summary(regression_pov_LFGCH4) #p-value: 5.653e-13

cor.test(svi_sf_join$WWTPCH4t,svi_sf_join$E_POV) #0.97
regression_pov_WWTPCH4<- lm(svi_sf_join$WWTPCH4t~svi_sf_join$E_POV)
summary(regression_pov_WWTPCH4) #p-value: < 2.2e-16

cor.test(svi_sf_join$OWCH4t,svi_sf_join$E_POV)#0.98
regression_pov_OWCH4<- lm(svi_sf_join$OWCH4t~svi_sf_join$E_POV)
summary(regression_pov_OWCH4) #p-value: < 2.2e-16

cor.test(svi_sf_join$AMCH4t,svi_sf_join$E_POV) #0.05
regression_pov_AMCH4<- lm(svi_sf_join$AMCH4t~svi_sf_join$E_POV)
summary(regression_pov_AMCH4) #p-value: 0.6652


```


```{r echo=FALSE, message=FALSE, warning=FALSE}
Plot_poor_CH4 <- 
  ggplot(svi_sf_join, aes(x = TotalCH4t, y = E_POV,)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 1000000)+
  labs(y= "Population Density", x = "Methane Generation Potential")+
  ggtitle("Total Methane Generation Potential v.s. Poor Population in California")+
  mytheme
  print(Plot_poor_CH4) 
  
Plot_poor_LF <- 
  ggplot(svi_sf_join, aes(x = LFGCH4t, y = E_POV,)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 750000)+
  labs(y= "Population Density", x = "Methane Generation Potential")+
  ggtitle("Landfill CH4 Generation Potential v.s. Poor Population in California")+
  mytheme
  print(Plot_poor_LF) 

Plot_poor_WW <- 
  ggplot(svi_sf_join, aes(x = WWTPCH4t, y = E_POV,)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 30000) +
  ylim(0, 750000)+
  labs(y= "Population Density", x = "Methane Generation Potential")+
  ggtitle("Wastewater CH4 Potential v.s. Poor Population in California")+
  mytheme
  print(Plot_poor_WW) 
  
Plot_poor_OW <- 
  ggplot(svi_sf_join, aes(x = OWCH4t, y = E_POV,)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 30000) +
  ylim(0, 750000)+
  labs(y= "Population Density", x = "Methane Generation Potential")+
  ggtitle("Organic Waste CH4 Potential v.s. Poor Population in California")+
  mytheme
  print(Plot_poor_OW)
  
Plot_poor_AM <- 
  ggplot(svi_sf_join, aes(x = AMCH4t, y = E_POV,)) +
  geom_point(alpha=2, color = "Purple") +
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 110000) +
  ylim(0, 400000)+
  labs(y= "Population Density", x = "Methane Generation Potential")+
  ggtitle("Manure CH4 Potential v.s. Poor Population in California")+
  mytheme
print(Plot_poor_AM)

```

\newpage

# Conclusion
The null hypothesis to our research question is that there is no correlation between population density and biogas generation potential. We ran a correlation test to test the strength and direction between California population and the total biogas generation potential.  We also ran a simple linear regression model to investigate whether these two variables involves a linear relationship. By looking at the residual plots, we can tell that the regression model fits our data but has outliers that disrupt the model accuracy. The p-value is 4.22e-15, indicating that the result is statistically significant. The multiple R-squared is 0.67, meaning that the independent variables can expalin 67% of the variations in biogas generation potential. Therefore, we are confident in rejecting the null hypothesis. Now we can say the total biogas generation potential in california does have a linear relationship with the population density. However, we still want to know how and what different sources is correlated with population. Subsequently, we ran the same test for each individual biogas sources. Based on the results from running correlation test and linear regression model for each of the biogas generation sources: Organic waste, Animal manure, Wastewater, and Landfill. We found out the the organic waste and wastewater contributes to the highest correlation to the population density, while Animal manure and landfill are less correlated. 

Overall, the analysis conducted show that biogas generation potential across 
counties in California is positively correlated with the population density. 
This implies that the higher the population in a specific area, the higher the 
biogas generation potential. However, the analysis also recognises that 
generation potential and correlation with population density varies according to
the biomass source across counties. As such, results show that animal based 
methane appeared to be less correlated with population. Assumptions that best 
explain this result are that more animals will contribute to animal-based 
biomass and that human habitation is often less where animal population is 
dense. Results of our analysis also show that the potential of landfill-based 
biogas was not limited to densely or sparsely populated areas, while organic 
waste and waste water represented the most significant sources of biogas 
generation potential across California. 

Future analysis can focus on improving the model presented in this report by 
eliminating some outliers and using more recent datasets. This study made use of
older data sets because those were the ones most recently made available on open
and credible data portals. Other areas of future analysis include exploring the 
level and process of utilization of California's biogas generation potential, 
as well as how utilization can be scaled. For example, the potential of 
leveraging the biogas generation potential by enabling counties with lower 
generation potentials to tap into the resources for those with higher 
generation.This also includes scoping the potential to utilize biogas energy 
for base load power or peaking, using various analyses including a cost-benefit 
analysis.



\newpage

# References

\<add references here if relevant, otherwise delete this section>
